\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-06-23 coqimpl.tex
Reply part of the comments.

Despite possible difficulties for future maintenance, this approach has
various advantages: 1. it is the easiest and the most accessible way to
prototype as it relieves us the necessity of familarity of Coq base,
especially for the implementation of the module and functor; 2. we have
a clear definition of trusted-base---the whole Coq; 3. it is easy to
debug---we just need to check the translated commands; 4. it can be
well-incorporated with the existent tools like VSCoq; 5. it is more
accessible for the interested audience who can then easily adapt our
plugin and give a try---otherwise they have to download and re-build the
whole customized Coq; 6. it is more stable because the surface syntax of
Coq should be stable across different versions; 7. still, this plugin
can capture the key ingredients of implementing family polymorphism
inside Coq and act as a reference for guiding an appropriate
implementation of Family Polymorphism inside all sorts of proof
assistants.\YZ{Does the Coq implementation perform type checking to
prevent illegal uses fam poly?}

\textbf{Family compiled into module.} This is the main idea of compiling
family components into Coq's primitives: we compile family into Coq
module, family type into module type, and the context of a given term
into parameters of the module. Unlike module, we need to faithfully
reflect the \textit{late binding} nature of family: for each defining
field $ .. \cL\sigma \vdash t : T $, we need to compile $t$ using
``universal-quantifier-wrapped term'' : $.. \vdash \lambda t : \forall
\cL \sigma. T$. However, instead of using Coq's universal quantifier, we
choose module and module parameters to achieve this wrapping---each
family type in the context of the judgement will be one module parameter
with the compiled module type, and thus each field of the family will
compile into a functor parametrized by its context. Doing so we can also
get rapid feedback from type checking when the users are defining each
field interactively with Coq.
% Insert one pseudo-code example for explanation of the mechanism
\begin{figure}[!htb]
  \begin{minipage}[t]{0.25\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Family A.
Field a : nat 
  := 1.
Field b 
  : self_A.a = self_A.a
  := eq_refl.
EndFamily.
\end{minted}
  \end{minipage}\YZ{Can the self's be omitted?}
  \begin{minipage}[t]{0.4\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Module a_4 (self_A: EmptySig).
Definition a : nat := 1. End a_4.
Module Type a_5 (self_A: EmptySig).
Parameter (a : nat). End a_5.
Module Type a_6 := a_5 EmptyMod.
\end{minted}
  \end{minipage}
  \begin{minipage}[t]{0.3\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Module b_7 (self_A: a_6).
Definition b 
  : self_A.a = self_A.a 
  := eq_refl. End b_7.

@@

% Concrete details of compilation of a family term data structure
Taking \cref{fig:plugin-example1} as an example: when the user
interactively inputs the line "Field a : nat := ..", our plugin will
translate this statement into the "Module a_9", "Module Type a_10, a_11"
three components.\YZ{Typo? Are they a_4, a_5, a_6 in Figure 2?}
{The "EmptySig" is just empty module type.} Then,
these modules and module types will be generated and type-checked by Coq
immediately. The computed "Module a_10" is then part of the context of
"Field b" (as the module type of "self__A") for "b" to refer to. The
internal representation of family "A" is simply the list of these
``compiled'' modules and module types. We can simply achieve incremental
checking and overriding as in \ref{langdesign-req0}---for example, if we
want to override "Field a", we simply replace that compiled module with
the new module inside the list. This compiled module type serves as an
abstraction that enables overriding.

Now that we know a family data structure is actually a list of compiled
modules, we may want to aggregate them into one complete compiled
module. We achieve that by repeatedly using Coq's "Include" command as
the example illustrated. Note that, in our example, we include two
functors without arguments---this is due to a perculiarity of Coq's
"Include": non-instantiated parameters will be automatically
instantiated by the appropriate fields in the surrounding defining
module.\YZ{'For example, ...'}
The resulting module is just like mundane Coq's module and thus
enjoy all the computational mechanism from Coq for free.

@@

\textbf{Reasoning Requires Non-Overridable.} This compilation strategy
can allow fields to be type-dependent on former fields thus fulfilling
\ref{langdesign-req1}. However, there are some caveats, especially when
the former fields are actually ``proof-relevant''. For example, we may
define $"add" : \mathbb{Z} \to \mathbb{Z} \to \mathbb{Z}$ as one field
and intend to prove its commutativity in a following field $"comm"$.
But to prove so, we need to know more than just a type interface of
"add" but also its \textbf{concrete definition}. Otherwise, in the
future, once "add" is overridden into a subtraction function then
"comm" will have problem to be inherited.
A more concrete example is in \cref{fig:plugin-example1}, consider the
case we have "Field b : self_A.a = 1 := .." instead. Then our
compilation and Coq's type checking will both fail: in the context of
"b" (module type "a_6") we only know "a" is of type "nat" but nothing
more (and this is necessary so that "a" can be overridden in the
future). To fix that, we need more information on "a" than its type---we
need to expose the concrete definition "a = 1" to the context of "b".
\YZ{I had a hard time working through these two paras. I guess they are
saying that there is a need to expose definitional equalities, and
therefore we need non-overridables. This causal flow should be made
more explicit.}

To achieve such exposure of definition, in meta-theory we only need to
expose the information as propositional equality as part of the
signature, and in our Coq plugin, we can achieve that even easier by
exposing the whole definition directly into module type. Doing so we can
have a better judgemental equality and we prohibit overriding because
subtraction doesn't have the same definition as $add$. We require the
users to decorate field as `Final' so that the plugin will proceed with
this special treatment.\YZ{This para is confusing: which kind of equality is used, propositional or judgmental? Where does 'Final' show up in the example?}

>>>>

Despite possible difficulties for future maintenance, this approach has
various advantages: 1. it is the easiest and the most accessible way to
prototype as it relieves us the necessity of familarity of Coq base,
especially for the implementation of the module and functor; 2. we have
a clear definition of trusted-base---the whole Coq; 3. it is easy to
debug---we just need to check the translated commands; 4. it can be
well-incorporated with the existent tools like VSCoq; 5. it is more
accessible for the interested audience who can then easily adapt our
plugin and give a try---otherwise they have to download and re-build the
whole customized Coq; 6. it is more stable because the surface syntax of
Coq should be stable across different versions; 7. still, this plugin
can capture the key ingredients of implementing family polymorphism
inside Coq and act as a reference for guiding an appropriate
implementation of Family Polymorphism inside all sorts of proof
assistants.\YZ{Does the Coq implementation perform type checking to
prevent illegal uses fam poly?}\EDJreply{Yes. Because I define feature of fampoly as a direct translation into Coq feature.  The fampoly is like a shorthand into multiple Coq commands. \\
Do you find translation from my current metatheory into MLTT (without linkage) helpful? \\
Actually I am also not sure what this question is aiming towards. \\
Are you complaining there is no formal specification of my fampoly feature in paper and thus there is no proof my Coq surface syntax is related to metatheory? If so, I thought our Coq plugin is only advertising how powerful Fampoly is to solve expression problem in mechanized proving. I can decouple the relationship between our plugin and our metatheory if you find the gap is really there... And I can only claim metatheory is about a partial result on incorporating fampoly with dependent type, Coq plugin is using two example to illustrate how mechanized proving can get benefit and our Coq plugin is only inspired by our metatheory (and not claim any relationship). Doing so would require rearrange the metatheory section before this section. \\
But I don't think the gap is there at all because if you want all possible gaps to disappear, then the action of using debruijn is wrong because we need to prove the one using debruijn and the surface syntax that is not using debruijn are ``equivalent''. What's more, module is not sigma type because Module is not a term in Coq and thus there is no formalism of Module at all. Let alone I don't see the plausibility of mechanizing dependent type without debruijn. \\ 
Please comment back so that I know what corresponding modification is needed on the paper. 
}

\textbf{Family compiled into module.} This is the main idea of compiling
family components into Coq's primitives: we compile family into Coq
module, family type into module type, and the context of a given term
into parameters of the module. Unlike module, we need to faithfully
reflect the \textit{late binding} nature of family: for each defining
field $ .. \cL\sigma \vdash t : T $, we need to compile $t$ using
``universal-quantifier-wrapped term'' : $.. \vdash \lambda t : \forall
\cL \sigma. T$. However, instead of using Coq's universal quantifier, we
choose module and module parameters to achieve this wrapping---each
family type in the context of the judgement will be one module parameter
with the compiled module type, and thus each field of the family will
compile into a functor parametrized by its context. Doing so we can also
get rapid feedback from type checking when the users are defining each
field interactively with Coq.
% Insert one pseudo-code example for explanation of the mechanism
\begin{figure}[!htb]
  \begin{minipage}[t]{0.25\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Family A.
Field a : nat 
  := 1.
Field b 
  : self_A.a = self_A.a
  := eq_refl.
EndFamily.
\end{minted}
  \end{minipage}\YZ{Can the self's be omitted?}\EDJreply{Yes. But in our current plugin implementation it is not omitted. I think altering example is fine. If you think some small distance between example and the current implementation is fine, please comment back and I will remove all the self_}
  \begin{minipage}[t]{0.4\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Module a_4 (self_A: EmptySig).
Definition a : nat := 1. End a_4.
Module Type a_5 (self_A: EmptySig).
Parameter (a : nat). End a_5.
Module Type a_6 := a_5 EmptyMod.
\end{minted}
  \end{minipage}
  \begin{minipage}[t]{0.3\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Module b_7 (self_A: a_6).
Definition b 
  : self_A.a = self_A.a 
  := eq_refl. End b_7.

@@

% Concrete details of compilation of a family term data structure
Taking \cref{fig:plugin-example1} as an example: when the user
interactively inputs the line "Field a : nat := ..", our plugin will
translate this statement into the "Module a_4", "Module Type a_5, a_6"
three components.\YZ{Typo? Are they a_4, a_5, a_6 in Figure 2?}\EDJreply{Thanks, fixed.}
{The "EmptySig" is just empty module type.} Then,
these modules and module types will be generated and type-checked by Coq
immediately. The computed "Module a_10" is then part of the context of
"Field b" (as the module type of "self__A") for "b" to refer to. The
internal representation of family "A" is simply the list of these
``compiled'' modules and module types. We can simply achieve incremental
checking and overriding as in \ref{langdesign-req0}---for example, if we
want to override "Field a", we simply replace that compiled module with
the new module inside the list. This compiled module type serves as an
abstraction that enables overriding.

Now that we know a family data structure is actually a list of compiled
modules, we may want to aggregate them into one complete compiled
module. We achieve that by repeatedly using Coq's "Include" command as
the example illustrated. Note that, in our example, we include two
functors without arguments---this is due to a perculiarity of Coq's
"Include": non-instantiated parameters will be automatically
instantiated by the appropriate fields in the surrounding defining
module.\YZ{'For example, ...'}\EDJreply{I don't think explaining a Coq feature inside our paper is a good idea? But I have added this in the following line, please check and retain it if you find this ok. }
For example, "b_7" requires a module of type "a_6". Inside module "A", when "Include a_4" is done, we will have a field \mintinline{Coq}{a : nat} inside the current surrounding "A". Then when we \mintinline{Coq}{Include b_7} without specifying the module parameters, Coq will try to find fields in the surrounding module, (i.e. module "A") to instantiate \mintinline{Coq}{self_A: a_6}. This is satisfied by the \mintinline{Coq}{a : nat} that is included earlier.
The resulting module is just like mundane Coq's module and thus
enjoy all the computational mechanism from Coq for free.

@@

\textbf{Reasoning Requires Non-Overridable.} This compilation strategy
can allow fields to be type-dependent on former fields thus fulfilling
\ref{langdesign-req1}. However, there are some cases we need to expose the concrete definition of a given fields (especially when we are proving properties about it).

For example, we may
define $"add" : \mathbb{Z} \to \mathbb{Z} \to \mathbb{Z}$ as one field
and intend to prove its commutativity in a following field $"comm"$.
But to prove so, we need to know more than just a type interface of
"add" but also its \textbf{concrete fixpoint definition}. Otherwise, in the
future, once "add" is overridden into a subtraction function then
"comm" will have problem to be inherited.
A more concrete example is in \cref{fig:plugin-example1}, consider the
case we have "Field b : self_A.a = 1 := .." instead. Then our
compilation and Coq's type checking will both fail: in the context of
"b" (module type "a_6") we only know "a" is of type "nat" but nothing
more (and this is necessary so that "a" can be overridden in the
future).

To fix that, we need more information on "a" than its type---we
need to expose the concrete definition "a = 1" to the context of "b".
\YZ{I had a hard time working through these two paras. I guess they are
saying that there is a need to expose definitional equalities, and
therefore we need non-overridables. This causal flow should be made
more explicit.}\EDJreply{Good suggestion. I try to add this casual flow into the first line of the paragraph.}

In our Coq plugin, we can achieve such exposure of definition by
exposing the whole definition directly into module type. Doing so we also prohibit overriding because
subtraction doesn't have the same definition as $add$. We require the
users to decorate field as `Final' so that the plugin will proceed with
this special treatment.\YZ{This para is confusing: which kind of equality is used, propositional or judgmental? Where does 'Final' show up in the example?}\EDJreply{I rephrase it and remove the part on taking about meta-theory. Now everything is about judgemental so I don't have to emphasize judgemental. I wanted to make the example small but I guess I can add one `Final' statement into our example. }
